---
title: "Final Project Submission"
author: "Ale Cuevas [Add y'all names here]"
date: "12/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup}
library(tidyverse)
library(modelr)
library(runner)
source("cleaning.R")
set.seed(42)
```
## Question 
Are there racial inequalities in Covid rates in the US?

ANSWER: Despite inconsistencies in the data, there are clear major racial inequities due to Covid.

##Background

#Uncertainty

##Case Uncertainty 
```{r Case Data by Race Inconsistencies, echo=FALSE}
plot(pressure)
```

## Death Uncertainty
```{r Death Data by Race Uncertainity, echo=FALSE}
plot(pressure)
```

## Test Uncertainty
```{r Test Data by Race Uncertainity, echo=FALSE}
plot(pressure)
```

#Data

##Case Data
```{r Case Data by Race, echo=FALSE}
plot(pressure)
```

##Death Data
```{r Death Data by Race, echo=FALSE}
plot(pressure)
```

##Test Data
```{r Test Data by Race, echo=FALSE}
plot(pressure)
```

#Conclusions

 More consistent race reporting is needed to better understand racial impact.
 
 States should examine their situation individually and develop programs to alleviate inequality.

Overall we need to implement better safety/testing programs to reduce inequality
focus on lessening inequality
(confounding variable is economic status)
(both in being essential workers and in ability to get tested/decision to go to hospital)
